Industrial, large-scale model predictive control with structured neural networks

### 2021-02-11 paper.tex
sync a few files from master.

\begin{keyword}
Model predictive control \sep machine learning \sep deep neural networks \sep
large-scale problems
\end{keyword}

@@

\subsection{Structured neural network}
An intuitive approach is to build a feedforward NN that takes the triple ($x,
x_s, u_s$) as its input and outputs a control that is close to the optimal
control. This strategy has been proposed in \cite*{karg:lucia:2020} and 
\cite*{chen:saulnier:atanasov:lee:kumar:pappas:morari:2018}
for a single fixed steady state. The approach has not been demonstrated to scale
to the size of model dimensions and the setpoint tracking offset-free MPC
problem considered in this article. An issue with this strategy is that the NN
has no knowledge about the MPC feedback law, and can require large amounts of
training data in large state dimensions to obtain a reasonable controller. At a
minimum, the MPC feedback law is such that at the origin, the control action
taken by the MPC controller is zero. For the setpoint tracking MPC problem
considered in this article, the NN control law must be such that at steady-state
operations, the NN maintains the plant at the desired steady state such that the
controlled measurements are at their setpoints. This structure implies $u_s =
\kappa_{NN}(x = x_s, x_s, u_s)$, in which we use $\kappa_{NN}(\cdot)$ to denote
the NN control law. To incorporate this information in a NN, we introduce the
following architecture
\begin{align}
z_0 &= \begin{bmatrix}
    x', & x_s', & u_s', & x_s', & x_s', & u_s' \end{bmatrix}'\notag \\
    f_i(z_{i-1}) &= \begin{bmatrix}
        W_i & 0 \notag \\
        0 & W_i
          \end{bmatrix}z_{i-1}  + \begin{bmatrix}
            b_i \\
            b_i
            \end{bmatrix} \notag\\
    z_i &= g(f_i(z_{i-1})), \textnormal{for} \ i \in \bbI_{1:h} \notag \\
    u &= u_s + \begin{bmatrix}
  W_{h + 1}, & -W_{h + 1} \\
    \end{bmatrix} z_h
\label{eq:NNstruct}
\end{align}
in which $g(a) = \textnormal{max} (0, a)$, is the ReLU operation applied
element-wise on its vector input, $z_i$ is the output of each hidden layer of
the network, $z_0$ is the input to the network, $h$ is the number of hidden
layers, and $i \in \bbI_{1:h}$. The parameters to be optimized using training
data are ($W_i, b_i$) and $W_{h+1}$, which are matrices of appropriate
dimensions. At steady state, when $x=x_s$ in $z_0$, \eqref{eq:NNstruct} outputs
$u_s$ regardless of the choice of weights as the output of all hidden layers
($z_i$) have same subvector repeated in the upper and lower half and
$\begin{bmatrix} W_{h + 1}, & -W_{h+ 1} \\
      \end{bmatrix} z_h$ equals zero. The software tensorflow
\citep{abadi:agarwal:barham:brevdo:et-al:2015} is used for NN training purposes
in this article, and the structured architecture can be built with symbolic
differentiation tools available in tensorflow. When a rate-of-change penalty on
the control is included in the regulator QP, the input to the structured NN is
modified to 
\begin{align*}
    z_0 &= 
    \begin{bmatrix}
    x', & u_{-1}', & x_s', & u_s', & x_s', & u_s', & x_s', & u_s' \\
    \end{bmatrix}'
\end{align*}
in which $u_{-1}$ is the previous control applied to the plant and the
dimensions of the matrices ($W_i, b_i$) and $W_{h+1}$ are appropriately
adjusted. Even after training, the control produced by the NN may not satisfy
the hard actuator constraints, and the saturation function
($\textnormal{sat}(\cdot, \underline{u}, \overline{u})$) is applied to the
output of the NN.

@@

\begin{figure*}[!h]
    \centering
	\includegraphics[page=1, width=0.7\textwidth,
		height=0.5\textheight]{cstrs_comparision_plots.pdf}    
        \caption{Transient closed-loop actuator trajectories of the CSTRs with
        flash separator example with the NN and optimal MPC controllers. The
        performance of the NN and optimal MPC controllers are almost
        indistinguishable.}   
    \label{fig:cl_cstrs_inputs}
\end{figure*}

\begin{figure*}[!h]
    \centering
	\includegraphics[page=2, width=0.7\textwidth,
		height=0.5\textheight]{cstrs_comparision_plots.pdf} \caption{Transient
		closed-loop controlled measurement trajectories of the CSTRs with flash
		separator example with the NN and optimal MPC controllers. The
		performance of the NN and optimal MPC controllers are almost
		indistinguishable.}
		\label{fig:cl_cstrs_outputs}
\end{figure*}

@@

The related work by \cite{hertneck:kohler:trimpe:allgower:2018} examines the
robustness of nonlinear systems in feedback with approximate MPC controllers.
The MPC control law approximation error is treated as an input disturbance, and
a robust MPC formulation based on constraint tightening procedures is used to
establish the allowable approximation error. The robustness of NN controllers
trained with samples of the nominal MPC control law in presence of process
disturbances and state estimation errors has not been established to date in the
literature.

@@

The validation simulations are performed directly with the plant model for the
case studies in this section. For industrial deployment, the NN training should
be followed by an offline validation step of the trained network to verify the
quality of the NN controller before online deployment. This validation can be
performed by: a) quantifying the MSE of the trained NN controller on test data
generated using the optimal MPC, b) examining closed-loop performance metrics in
simulations carried out with the linear model used in MPC and the optimal MPC
and NN controllers. Additional probabilistic validation techniques for NNs
trained using robust MPC solutions for nonlinear problems are proposed in
\cite{karg:alamo:lucia:2019} and \cite{hertneck:kohler:trimpe:allgower:2018}.

@@

\begin{itemize}
    \item Controller performance index
    \begin{align*}
    \Lambda_k = \dfrac{1}{k}\sum_{t=1}^{k} & \big(\norm{x(t) - x_{s} (t)}^2_{Q} \\
        & + \norm{u(t) - u_{s} (t)}^2_{R} + \norm{\Delta u(t)}^2_{S}\big)         
    \end{align*}
    \item \% Performance loss = $100(\Lambda^{\textnormal{F}}_{N_{t}} -
    \Lambda^{\textnormal{MPC}}_{N_{t}})/\Lambda^{\textnormal{MPC}}_{N_{t}}$
    \item Average and worst-case speedups
    \item Data generation and NN training times
    \item Memory required to store the weights of the NNs
\end{itemize}
Here, $N_t$ is the number of simulation time steps,  
$\Lambda^{\textnormal{F}}_{N_{t}}$ and $\Lambda^{\textnormal{MPC}}_{N_{t}}$ are
the average stage costs obtained at the end of the simulation period by the fast
controllers and the MPC controller. The data generation and online timing
comparisons are performed on a computing cluster that has several multi-core
CPUs of clock speed 2.4GHz. The NN trainings are performed on a Tesla V100-SXM2
GPU that has 32 GB memory.

@@

\subsection{CSTRs with a flash separator}
For the first example, the two stirred tank reactors in series with a flash
separator depicted in Figure \ref{fig:schematic_cstrs} is considered as the
plant. In both the CSTRs, a desired product $B$ is produced with the reaction $A
\rightarrow B$, and an undesired product $C$ is produced with the reaction $B
\rightarrow C$. The reactant $A$ is the major component in the feed streams
supplied to the reactors. After the second CSTR, the reaction mixture is sent to
the non-adiabatic flash that separates the reactant $A$ from the product $B$.
The $A$ rich vapor phase is recycled back to the first CSTR. The plant is
simulated using the set of nonlinear ordinary differential equations (ODEs)
available in \cite*{venkat:2006}, Appendix. The model has 12 states ($H_r,
x_{Ar}, x_{Br}, T_r, H_m, x_{Am}, x_{Bm}, T_m, H_b, x_{Ab}, x_{Bb}, T_b$), 6
control inputs ($F_0, F_1, D, Q_r, Q_m, Q_b$), and 5 disturbances ($x_{A0},
x_{A1}, x_{B0}, x_{B1}, T_0$). The controlled measurements with setpoints are
the heights and temperatures of the three units, and we assume that all the
states are measured for this example. The sample time for the measurements is
chosen as 10 seconds. The parameter values used to simulate the ODEs for the
plant, the actuator constraints, and the bounds of the controlled measurements
and disturbances used for the data generation and validation simulations are
shown in Table \ref{table:cstrs_pars}, Appendix \ref{app:cstrs_pars}.

@@

First, the amount of data required to obtain an accurate NN controller is
studied. To this aim, NNs with the three chosen architectures are trained with
incremental increases of $10^4$ training samples starting from $4 \times 10^4$
to $1.5 \times 10^5$ samples. A batch size of 1024 samples is used in Adam, and
all the NNs are trained for 2000 epochs. For all the NN training performed,
10$\%$ of the training dataset is kept as a buffer dataset, which is monitored
during training and the network weights are updated only when the MSE decreases
on this buffer dataset after every training epoch. The NNs after training are
used in validation simulations with the setpoint and disturbance signals
generated for the expected plant behavior case. The closed-loop performance is
compared with the optimal MPC controller and the $\%$ performance loss obtained
is plotted in Figure \ref{fig:cstrs_statistics} (top left) as a function of the
number of training samples. All the trained NNs provide less than $1 \%$ loss in
this performance metric after training with about $9 \times 10^4$ samples. The
best $\%$ performance loss obtained by the NNs is summarized in Table
\ref{table:cstrs}.

@@

@@

The $\%$ performance loss obtained using the four heuristic controllers, their
corresponding speedups, maximum time required to train each NN, number of
parameters in each NN, and the memory required to store  
the weights of the NNs is summarized in the Table \ref{table:cstrs}. The $85 \%$
performance loss reported for the unstructured NN corresponds to training with
$1.5 \times 10^5$ samples and the performance is unacceptable for an industrial
application. The noticeable performance improvement with the structured NNs is
because the control law is exactly the same as the optimal feedback law at
steady-states, which simplifies the optimal feedback law approximation problem
since the training process only has to improve the NN control law for the
transient conditions.

@@

The key features for the scalability of the proposed approach are the structure
of the NN and data generation for only typical plant operational scenarios. The
NN design approach was demonstrated to scale on an industrial crude distillation
unit model with state and control dimensions of 252 and 32 respectively, and a
control-sample horizon length of 140. After the offline design phase, NNs
execute MPC about three to five orders of magnitude faster than an available QP
solver with less than 1$\%$ performance degradation in the closed-loop. The
deployment decision of the NNs must be based on the ability to reliably identify
and sample the anticipated plant operational scenarios with setpoints and large
magnitude disturbances that may change for a reasonable duration of operation.
Parallel computing can be used for the offline data generation step if solving
QPs for each sampled state starts to take excessive amounts of time. After the
training process, the NNs are easily deployable on memory constrained hardware.
We established conditions under which the NN controllers are robust to state
estimation errors and process disturbances, provides some theoretical support to
practitioners interested in the deployment of NNs. Future research may be
focused to establish conditions for guaranteed closed-loop stability with NN
controllers.

@@

\section*{Source code}
The code for the simulation studies performed in this article can be found at:
\url{https://github.com/pratyushkumar211/industrial_nnmpc_2021}

>>>>

\begin{keyword}
Model predictive control \sep machine learning \sep deep neural networks \sep
large-scale systems \sep quadratic programming \sep inherent robustness
\end{keyword}

@@

\subsection{Structured neural network}
An intuitive approach is to build a feedforward NN that takes the triple ($x,
x_s, u_s$) as its input and outputs a control that is close to the optimal
control. This strategy has been proposed in \cite*{karg:lucia:2020} and 
\cite*{chen:saulnier:atanasov:lee:kumar:pappas:morari:2018}
for a single fixed steady state. The approach has not been demonstrated to scale
to the size of model dimensions and the setpoint tracking offset-free MPC
problem considered in this article. An issue with this strategy is that the NN
has no knowledge about the MPC feedback law, and can require large amounts of
training data in large state dimensions to obtain a reasonable controller. At a
minimum, the MPC feedback law is such that at the origin, the control action
taken by the MPC controller is zero. For the setpoint tracking MPC problem
considered in this article, the NN control law must be such that at steady-state
operations, the NN maintains the plant at the desired steady state such that the
controlled measurements are at their setpoints. This structure implies $u_s =
\kappa_{NN}(x = x_s, x_s, u_s)$, in which we use $\kappa_{NN}(\cdot)$ to denote
the NN control law. To incorporate this information in a NN, we introduce the
following architecture
\begin{align}
z_0 &= \begin{bmatrix}
    x', & x_s', & u_s', & x_s', & x_s', & u_s' \end{bmatrix}'\notag \\
    f_i(z_{i-1}) &= \begin{bmatrix}
        W_i & 0 \notag \\
        0 & W_i
          \end{bmatrix}z_{i-1}  + \begin{bmatrix}
            b_i \\
            b_i
            \end{bmatrix} \notag\\
    z_i &= g(f_i(z_{i-1})), \textnormal{for} \ i \in \bbI_{1:h} \notag \\
    u &= u_s + \begin{bmatrix}
  W_{h + 1}, & -W_{h + 1} \\
    \end{bmatrix} z_h
\label{eq:NNstruct}
\end{align}
in which $g(a) = \textnormal{max} (0, a)$, is the ReLU operation applied
element-wise on its vector input, $z_i$ is the output of each hidden layer of
the network, $z_0$ is the input to the network, $h$ is the number of hidden
layers, and $i \in \bbI_{1:h}$. The parameters to be optimized using training
data are ($W_i, b_i$) and $W_{h+1}$, which are matrices of appropriate
dimensions. At steady state, when $x=x_s$ in $z_0$, \eqref{eq:NNstruct} outputs
$u_s$ regardless of the choice of weights and the activation function, as the
output of all hidden layers ($z_i$) have same subvector repeated in the upper
and lower half and the term $\begin{bmatrix} W_{h + 1}, & -W_{h+ 1} \\
      \end{bmatrix} z_h$ equals zero. The NN structure \eqref{eq:NNstruct} is
equivalent to the difference in outputs of two standard NNs with inputs ($x,
x_s, u_s$) and ($x_s, x_s, u_s$) respectively, with $u_s$ added to the final
output. The software tensorflow \citep{abadi:agarwal:barham:brevdo:et-al:2015}
is used for NN training purposes in this article, and the structured
architecture can be built with symbolic differentiation tools available in
tensorflow. When a rate-of-change penalty on the control is included in the
regulator QP, the input to the structured NN is modified to 
\begin{align*}
    z_0 &= 
    \begin{bmatrix}
    x', & u_{-1}', & x_s', & u_s', & x_s', & u_s', & x_s', & u_s' \\
    \end{bmatrix}'
\end{align*}
in which $u_{-1}$ is the previous control applied to the plant and the
dimensions of the matrices ($W_i, b_i$) and $W_{h+1}$ are appropriately
adjusted. Even after training, the control produced by the NN may not satisfy
the hard actuator constraints, and the saturation function
($\textnormal{sat}(\cdot, \underline{u}, \overline{u})$) is applied to the
output of the NN.

@@

\begin{figure*}[!h]
    \centering
	\includegraphics[page=1, width=0.7\textwidth,
		height=0.46\textheight]{cstrs_comparision_plots.pdf}    
        \caption{Transient closed-loop actuator trajectories of the CSTRs with
        flash separator example with the NN and optimal MPC controllers. The
        performance of the NN and optimal MPC controllers are almost
        indistinguishable.}   
    \label{fig:cl_cstrs_inputs}
\end{figure*}

\begin{figure*}[!h]
    \centering
	\includegraphics[page=2, width=0.7\textwidth,
		height=0.46\textheight]{cstrs_comparision_plots.pdf} \caption{Transient
		closed-loop controlled measurement trajectories of the CSTRs with flash
		separator example with the NN and optimal MPC controllers. The
		performance of the NN and optimal MPC controllers are almost
		indistinguishable.}
		\label{fig:cl_cstrs_outputs}
\end{figure*}

@@

The related work by \cite{hertneck:kohler:trimpe:allgower:2018} examines the
robustness of nonlinear systems in feedback with approximate MPC controllers.
The MPC control law approximation error is treated as an input disturbance, and
a robust MPC formulation based on constraint tightening procedures is used to
establish the allowable approximation error. The robustness of NN controllers
trained with samples of nominal MPC control laws in presence of process
disturbances and state estimation errors has not been established to date in the
literature.

@@

The validation simulations are performed directly with the plant model for the
case studies in this section. For industrial deployment, the NN training should
be followed by an offline validation step of the trained network to verify the
quality of the NN controller before online deployment. This validation can be
performed by: a) quantifying the MSE of the trained NN controller on test data
generated using the optimal MPC controller, b) examining closed-loop performance
metrics in simulations carried out with the linear model used in MPC and the
optimal MPC and NN controllers, similar to the simulations performed in this
section with the plant model. Additional probabilistic validation techniques for
NNs trained using robust MPC solutions for nonlinear models are proposed in
\cite{karg:alamo:lucia:2019} and \cite{hertneck:kohler:trimpe:allgower:2018}.

@@

\begin{itemize}
    \item Controller performance index
    \begin{align*}
    \Lambda_k = \dfrac{1}{k}\sum_{t=1}^{k} & \big(\norm{x(t) - x_{s} (t)}^2_{Q} \\
        & + \norm{u(t) - u_{s} (t)}^2_{R} + \norm{\Delta u(t)}^2_{S}\big)         
    \end{align*}
    \item \% Performance loss = $100(\Lambda^{\textnormal{F}}_{N_{t}} -
    \Lambda^{\textnormal{MPC}}_{N_{t}})/\Lambda^{\textnormal{MPC}}_{N_{t}}$
    \item Average and worst-case speedups
    \item Data generation and NN training times
    \item Memory required to store the weights of the NNs
\end{itemize}
Here, $N_t$ is the number of simulation time steps,  
$\Lambda^{\textnormal{F}}_{N_{t}}$ and $\Lambda^{\textnormal{MPC}}_{N_{t}}$ are
the average stage costs obtained at the end of the simulation period by the fast
controllers and the MPC controller. The data generation and online timing
comparisons are performed on a computing cluster that has several multi-core
CPUs of clock speed 2.4GHz. The NN trainings are performed on a Tesla V100-SXM2
GPU that has 32 GB memory. The code for the simulation studies are available at:
\url{https://github.com/pratyushkumar211/industrial_nnmpc_2021}.

@@

\subsection{CSTRs with a flash separator}

For the first example, the two stirred tank reactors in series with a flash
separator depicted in Figure \ref{fig:schematic_cstrs} is considered as the
plant. In both the CSTRs, a desired product $B$ is produced with the reaction $A
\rightarrow B$, and an undesired product $C$ is produced with the reaction $B
\rightarrow C$. The reactant $A$ is the major component in the feed streams
supplied to the reactors. After the second CSTR, the reaction mixture is sent to
the non-adiabatic flash that separates the reactant $A$ from the product $B$.
The $A$ rich vapor phase is recycled back to the first CSTR. The plant is
simulated using the set of nonlinear ordinary differential equations (ODEs)
available in \cite*{venkat:2006}, Appendix. The model has 12 states ($H_r,
x_{Ar}, x_{Br}, T_r, H_m, x_{Am}, x_{Bm}, T_m, H_b, x_{Ab}, x_{Bb}, T_b$), 6
control inputs ($F_0, F_1, D, Q_r, Q_m, Q_b$), and 5 disturbances ($x_{A0},
x_{A1}, x_{B0}, x_{B1}, T_0$). The controlled measurements with setpoints are
the heights and temperatures of the three units, and we assume that all the
states are measured for this example. The sample time for the measurements is
chosen as 10 seconds. The parameter values used to simulate the ODEs for the
plant, the actuator constraints, and the bounds of the controlled measurements
and disturbances used for the data generation and validation simulations are
shown in Table \ref{table:cstrs_pars}, Appendix \ref{app:cstrs_pars}.

@@

@@

First, the amount of data required to obtain an accurate NN controller is
studied. To this aim, NNs with the three chosen architectures are trained with
incremental increases of $10^4$ training samples starting from $4 \times 10^4$
to $1.5 \times 10^5$ samples. A batch size of 1024 samples is used in Adam, and
all the NNs are trained for 2000 epochs. For all the NN trainings performed,
10$\%$ of the training dataset is kept as a buffer dataset, which is monitored
during training and the network weights are updated only when the MSE decreases
on this buffer dataset after every training epoch. The NNs after training are
used in validation simulations with the setpoint and disturbance signals
generated for the expected plant behavior case. The closed-loop performance is
compared with the optimal MPC controller and the $\%$ performance loss obtained
is plotted in Figure \ref{fig:cstrs_statistics} (top left) as a function of the
number of training samples. All the trained NNs provide less than $1 \%$ loss in
this performance metric after training with about $9 \times 10^4$ samples. The
best $\%$ performance loss obtained by the NNs is summarized in Table
\ref{table:cstrs}.

@@

The $\%$ performance loss obtained using the four heuristic controllers, their
corresponding speedups, maximum time required to train each NN, number of
parameters in each NN, and the memory required to store the weights of the NNs
is summarized in the Table \ref{table:cstrs}. The memory requirement shows that
NNs are deployable on resource constrained hardware, which can be crucial for
industrial deployment at regulatory control layers. The $85 \%$ performance loss
reported for the unstructured NN corresponds to training with $1.5 \times 10^5$
samples and the performance is unacceptable for an industrial application. The
noticeable performance improvement with the structured NNs is because the
control law is exactly the same as the optimal feedback law at steady-states,
which simplifies the optimal feedback law approximation problem since the
training process only has to improve the NN control law for the transient
conditions.

@@

The key features for the scalability of the proposed approach are the structure
of the NN and data generation for only typical plant operational scenarios. The
NN design approach was demonstrated to scale on an industrial crude distillation
unit model with state and control dimensions of 252 and 32 respectively, and a
control-sample horizon length of 140. After the offline design phase, NNs
execute MPC about three to five orders of magnitude faster than an available QP
solver with less than 1$\%$ performance degradation in the closed-loop. The
deployment decision of the NNs must be based on the ability to reliably identify
and sample the anticipated plant operational scenarios with setpoints and large
magnitude disturbances that may change for a reasonable duration of operation.
Parallel computing can be used for the offline data generation step if solving
QPs for each sampled state starts to take excessive amounts of time. After the
training process, the NNs are easily deployable on memory constrained hardware.
We established conditions under which the NN controllers are robust to state
estimation errors and process disturbances, provides some theoretical support to
practitioners interested in the deployment of NNs.

The online QP based MPC was adopted by practitioners due to its ability to
systematically handle multivariable systems, process constraints, improvements
in QP solution algorithms, and has guaranteed stability properties. The
additional computing hardware for training in large applications and sampling
the plant operational scenarios are additional complexities required to design
NN controllers. The advantage of NNs is their speed of MPC controller execution
and convenience for deployment on memory constrained hardware. Future research
may be focused to establish conditions for guaranteed closed-loop stability with
NN controllers, systematically handling state constraints, approximating the
target selector QP, and extensions to large, nonlinear MPC applications.

@@
